\section{Vocabularies}
A vocabulary with name {\tt MyVoc} is declared by
\begin{lstlisting}
	vocabulary MyVoc {
		// contents of the vocabulary
	}
\end{lstlisting}
A vocabulary can contain symbol declarations, symbol pointers, and other vocabularies. Symbols are types (sorts), predicate and functions symbols.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Symbol declarations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\label{ssec:symbols}
A type with name \code{MyType} is declared by
\begin{lstlisting}
	type MyType
\end{lstlisting}
When declaring a type, it can be stated that this type is a subtype or supertype of a set of other types.  The following declares \code{MyType} to be a subtype of the previously declared types \code{A1} and \code{A2}, and a supertype of the previously declared types \code{B1} and {B2}:
\begin{lstlisting}
	type MyType isa A1, A2 contains B1, B2
\end{lstlisting}
In the rest of this text, we will sometimes use ``parent type'' as direct supertype and ``ancestor type'' for (in)direct supertype.

Types can also be \emph{constructed} using the \code{constructed from} keywords. A constructed type has a fixed interpretation which is the union of the images of a set of \emph{constructor functions}. These constructor functions each have an interpretation which maps every tuple from their domain to a unique domain element in the interpretation of the constructed type. Constructor functions should not be explicitly declared by the vocabulary; a function occurring in the declaration is automatically declared as a constructor function. E.g., a constructed type \code{Square} with constructors \code{Index/2} and \code{SpecialSquare/0} is declared as by
\begin{lstlisting}
	type T1
	type Square constructed from {Position(T1,T1),SpecialSquare}
\end{lstlisting}
Note that constructed types currently do not support any type inheritance.\\

Also note that it is impossible to construct numerical types, since numerals simply are not constructor functions. However, sometimes it is useful to interpret a numerical type independent of any structure. This can be done in the vocabulary with a similar syntax as when the type would be interpreted in a structure. It is possible to make such a type inherit from a numerical supertype, or make it contain a numerical subtype.
\begin{lstlisting}
	type SmallPrimes = {2;3;5;7} isa nat 
\end{lstlisting}

A predicate with name \code{MyPred}, arity 3 and types \code{T1},\code{T2},\code{T3} is declared by
\begin{lstlisting}
	MyPred(T1,T2,T3) 
\end{lstlisting}

A predicate with arity zero can be declared by \code{MyPred()} or \code{MyPred}.

A function with name \code{MyFunc}, input types \code{T1},\code{T2},\code{T3} and output type \code{T}  is declared by
\begin{lstlisting}
	MyFunc(T1,T2,T3):T
\end{lstlisting}

A partial function is declared by 
\begin{lstlisting}
	partial MyFunc(T1,T2,T3):T
\end{lstlisting}

Constants of type \code{T} can be declared by \code{MyConst:T} or \code{MyConst():T}.

Any symbol has to be declared on its own line.

Invalid symbol names are: isa, type, extends, contains, extern



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Symbol pointers}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

To include a type, predicate, or function from a previously declared vocabulary \code{V} in another vocabulary \code{W}, write
\begin{lstlisting}
	/* Declaration of vocabulary V*/
	vocabulary V {
		//...
		type A
		P(A)
		F(A,A):A
		//...
	}

	vocabulary W {
		extern type V::A
		extern V::P[A]		//also possible: extern V::P/1
		extern V::F[A,A:A]	//also possible: extern V::F/2:1
	}
\end{lstlisting}

\todo{verwijder using en extern, zie meeting. Het wordt extends, right?}

In the example, explicitly including type \code{A} of vocabulary \code{V} in \code{W} is not needed, since types of included predicates or functions are automatically included themselves.  To include the whole vocabulary \code{V} in \code{W} at once, used
\begin{lstlisting}
	vocabulary W {
		extern vocabulary V
	}
\end{lstlisting}




%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The standard vocabulary}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The global namespace contains a fixed vocabulary \code{std}, which is defined as follows:

\begin{lstlisting}
	vocabulary std {
		type nat	
		type int contains nat

		+(int,int) : int
		-(int,int) : int
		*(int,int) : int
		partial /(int,int) : int
		%(int,int) : int
		abs(int) : int
		-(int) : int

	}
\end{lstlisting}
Every vocabulary implicitly contains all symbols of \code{std}.  Also, every vocabulary contains for each of its types \code{A} the predicates \code{=(A,A)}, \code{<(A,A)}, and \code{>(A,A)} and the functions \code{MIN:A}, \code{MAX:A}, \code{SUCC(A):A} and \code{PRED(A):A}. In every structure, the symbols of \code{std} have the following interpretation:

\begin{center}
	
\begin{tabular}{l|l}
\code{nat} & all natural numbers \\
\code{int} & all integer numbers \\ 
\code{+(int,int) : int} & integer addition \\
\code{-(int,int) : int } &integer subtraction \\
\code{*(int,int) : int} & integer multiplication \\
\code{/(int,int) : int} & division (only defined when the second argument is a non-zero divisor of the first) \\
\code{\%(int,int) : int} & remainder \\
\code{abs(int) : int} & absolute value \\
\code{-(int) : int} & unary minus \\
\end{tabular} 

\end{center}

The predicate \code{=/2} is always interpreted by equality.  The order $<_{dom}$ on domain elements is defined by
\begin{itemize}
	\item numbers are smaller than non-numbers;
%\item strings are smaller than compound domain elements (see below for a definitions of a compound domain element);
\item $d_1<_{dom} d_2$ if $d_1$ and $d_2$ are numbers and $d_1<d_2$;
\item $d_1<_{dom} d_2$ if $d_1$ and $d_2$ are strings that are not numbers and $d_1$ is before $d_2$ in the lexicographic ordering;
\item $d_1<_{dom} d_2$ is some total order on compound domain elements (which we do not specify).
\end{itemize}

Every structure contains the following fixed interpretations:

\begin{center}
	
\begin{tabular}{l|l}
\code{<(A,A)} & the projection of $<_{dom}$ to the domain of A \\
\code{>(A,A)} & the projection of $>_{dom}$ to the domain of A \\
\code{MIN:A} & the $<_{dom}$-least element in the domain of A \\
\code{MAX:A} & the $<_{dom}$-greatest element in the domain of A \\
\code{SUCC(A):A} & the partial function that maps an element a  of the domain of \code{A} \\ & to the $<_{dom}$-least element of the domain of A that is strictly larger than a \\
\code{PRED(A):A} & the partial function that maps an element a  of the domain of \code{A} \\ & to the $<_{dom}$-greatest element of the domain of A that is strictly smaller than a \\
\end{tabular} 

\end{center}

In an IDP-file, you should disambiguate which \code{MAX} you want to use.  This is done by \code{MAX[:MyType]}.



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{LTC vocabularies}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

You can declare a vocabulary to be an LTC-vocabulary, as defined \href{http://dtai.cs.kuleuven.be/krr/files/progression.pdf}{here}. Thus, a vocabulary containing a type $Time$, a constant $Start$ typed $Time$ and a function $Next$, $Time\to Time$. 
If these types and functions have the above name, you do this as follows
\begin{lstlisting}
	LTCvocabulary V {
		type Time	
		Start:Time
		Next(Time):Time
		...//Rest of your vocabulary
	}
\end{lstlisting} 

Doing so automatically creates  vocabularies $V\_ss$ (singlestate vocabulary) and $V\_bs$ (bistate vocabulary). 

In case your type $Time$ is called differently, you can inform IDP about this with:

\begin{lstlisting}
	LTCvocabulary V (MyTime, MyStart, MyNext) {
		type MyTime	
		MyStart:MyTime
		MyNext(MyTime):MyTime
		...//Rest of your vocabulary
	}
\end{lstlisting} 





